Human capital mediates natural selection in contemporary humans

### 2021-12-30 why-natural-selection.Rmd
Shorten the parents section; discuss mediation analysis; add TOC to appendix; explain choice of PGS.

---
title: "Human capital mediates natural selection in contemporary humans"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

options(digits = 2)
options(dplyr.summarise.inform = FALSE) # shut up about summarize
knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE,
        fig.height = 3.5
      )
options(huxtable.long_minus = TRUE)

huxtable::set_default_properties(latex_float = "h!")

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] Note that equation 
\eqref{eq:regression} does not include many possible environmental and
genetic confounds that could affect fertility, and as a result, $\beta$ is not
an estimate of the causal effect of a polygenic score on fertility. However, 
natural selection is a correlational phenomenon, not a causal one: if a 
polygenic score correlates with low fertility, it is being selected against, 
whatever the underlying causal relationship.

@@

@@

## Selection in the parents' generation

The UK Biobank data contains information on respondents' number of
siblings. Since respondents' polygenic scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has an additional source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Selection effects are highly correlated across the two generations, and most
share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Absolute effect
size estimates are larger for the parents' generation. We treat this result
cautiously, because when we split respondents up by year of birth, we find few
differences in effect sizes between early- and late-born respondents, for either
generation. In other words, since estimated effect sizes change across
"generations", but do not change over time within either generation, the change
may be due to remaining ascertainment bias within the sample or other effects.
In particular, effect sizes in both generations may depend on polygenic scores'
correlation with childlessness, and we cannot estimate this for the parents'
generation.

Although the direction of selection effects does not change between the
generations, there are other differences. Compared to our standard polygenic
scores (residualized on 100 principal components of genetic data), 
selection effects on unresidualized scores are about ten percent higher
on average for *number of siblings*, whereas effects for *number of children*
barely change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could
be because earlier fertility is driven more by geographically clustered
deprivation [e.g. via an insurance motive, @rendall1998old], which may correlate
with the broad-scale genetic variation captured by principal components, such as
ancestry.

We also check whether selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results (Appendix Figure
\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas.

Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
number of children and number of grandchildren (for the parents' generation). We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. A possible explanation for this comes 
from the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect leads higher earners to have more children. The
substitution effect leads them to have fewer children.

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is concave and increasing. We treat $N$ as continuous, in line with
the literature: this can be thought of as the expected number of children
among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

The above does not aim to provide a complete theory of fertility. Rather, it
shows that a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

@@

drake::loadd(res_all)
drake::loadd(res_age_flb)
drake::loadd(ashe_income)

famhist <- add_ashe_income(famhist, ashe_income)

cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
# TODO: use age_left_edu when this includes college

cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")

@@

dfr <- data.frame(
        PGS          = rep(score_names, 4),
        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. Substitution effects
dominate income effects overall, which fits the known association between income
and lower fertility [@becker1960economic;@jones2006economic]. The correlations
reverse when we control for age at first live birth, suggesting that within AFLB
categories, the income effect dominates. (See also the Supplementary Animations
for the uncorrected effects per AFLB group, which show the correlation reversing
after age 22.)

Second we test whether educational attainment mediates the correlation between
polygenic scores and fertility.

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. This need not imply
that a given polygenic score is causal of either fertility or human capital. Instead
scores may simply correlate with causally relevant environments, for example via
"genetic nurture" effects [@kong2018nature] or population stratification. To 
investigate the presence of these indirect genetic effects, we estimate equation
\@ref(eq:regression) among full siblings, controlling for a full set of sibling
group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
sample size, all within-sibling effects are insignificant after Bonferroni
correction. However, effect sizes are positively correlated with effect sizes
from the pooled model, and about 1/3 smaller (regressing within-sibling on
pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
with the ~50% decrease in heritability in within-sibling GWASs on age at first
birth and educational attainment [@Howe2021.03.05.433935]. We see these results
as providing weak evidence that polygenic scores cause fertility, with effects
being at least partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores, effect sizes are reduced by about 20%.

@@

We consider three alternative theories that might explain our results. The first
is that welfare benefits which incentivize child-bearing might be taken up
among low-income individuals. However, our effect sizes appear unchanged
over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
\@ref(tab:tbl-children-by-period)), during which
government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter $a$
in the model [cf. @jones2008fertility]. This theory would not explain why our
results are weaker at higher incomes and education levels.[^note] A third
alternative is that traits under selection are linked to externalizing behaviour
and risk-seeking. This might be partially captured by our risk-aversion
parameter $\sigma$, which affects fertility by changing the marginal utility of
income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk aversion measure, which is a single yes/no question.
However, the risk aversion measure *does* predict the overall number of
children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
33 regressions). Given that, and the statistical power we get from our sample
size, we believe that the non-result is real: while risk aversion does predict
fertility in the sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced. And we are unsure
whether genetic variants underlying other phenotypes will show a similar pattern
of natural selection to those studied here. In addition, genetic effects on
educational attainment have been shown to be inflated in population-based
samples as compared to within-family designs, likely because of indirect genetic 
effects, gene-environment correlations, and/or assortative mating 
[@lee2018gene; @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935]. 
In short, it is probably too early to tell whether modern natural selection has a
substantively important effect on the genetic make-up of the population.
Nevertheless, we note that selection effects on our measured polygenic scores
are still relatively small, even after reweighting to account for ascertainment
bias.

@@

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
income attributable to the "true" polygenic score for educational attainment as
`r r2_income_true_psea` among respondents. Counterfactually holding environmental
conditions constant, natural selection would raise this proportion to 
`r r2_income_true_psea_wt` among respondents' children. Similarly, since many
polygenic scores are predictive of disease risk, they could potentially increase
health inequalities. In general, the evolutionary history of anatomically modern
humans is related to disease risk [@benton2021influence]; understanding the role
of contemporary natural selection may aid in mapping the genetic architecture of
current health disparities.

@@

Polygenic scores were computed by summing the alleles across ~1.3 million 
genetic variants weighted by their effect sizes as estimated in 33 genome-wide 
association studies (GWASs) that excluded UK Biobank. To control for population 
stratification, we corrected the polygenic scores for 100 principal components (PCs).

@@

@@

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings*, i.e.
parents' number of children, on polygenic scores. By definition, members of the
parents' generation who had no children cannot be included in this data. For a
clean comparison with the respondents' generation, we rerun regressions on
*number of children* excluding those with no children, and show
results in the figure.

@@

The fact that childless people have such a strong effect on estimates makes it
hard to compare total effect sizes across generations. In particular, since the
parents' generation has a different distribution of numbers of children,
childless people may have had more or less effect in that generation. Another
issue is that we are in effect estimating parents' polygenic scores by the
scores of their children. This introduces some noise into our independent
variable, which might lead to errors-in-variables and bias coefficients towards
zero.

@@

drake::loadd(res_townsend_parents)
drake::loadd(famhist_townsend_71)
n_regs <- nrow(res_townsend_parents)

res_townsend_parents %>% 
  mutate(
    `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
    `Townsend quintile` = add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", 
                                  data = famhist_townsend_71)
  ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5
                  )

rm(famhist_townsend_71)
```

@@

n_positive <- sum(res_mediation$prop > 0)
n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
median_prop <- median(res_mediation$prop)

@@

Table \@ref(tab:tbl-res-mediation) shows results. `r n_pos_sig` scores are 
significantly greater than 0 at $p = 0.05/33$.

@@

Table \@ref(tab:tbl-inequality-effects) shows correlations between children's
polygenic scores and household income (UKB data field 738). Column "With
selection" uses respondents' scores weighted by Age/Qualification, times number
of children. Column "Without selection" uses scores weighted by
Age/Qualification only, i.e. the counterfactual correlation if all respondents
had the same number of children (equivalently, the correlation in the
respondents' own generation).

@@

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ dues not differ between the two sets of
weighted regressions.

The estimate above uses a pooled regression of income on EA3, which may include
confounding environmental variation such as genetic nurture. Within-siblings
regression would give a smaller estimate of the effect of EA3. However, this
would also affect the reliability of the regression [@griliches1986errors],
making it hard to back out estimates of the true $R^2$. Nevertheless, we
can calculate the proportional increase in $R^2$ due to natural selection, again 
so long as the reliability coefficient is unaffected by
the weighting scheme. Running within-siblings regressions of EA3 on income group
gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this procedure estimates the proportional change in $R^2$ 
as `r prop_wt_fe`.

@@

# Model proofs

@@

\clearpage
\FloatBarrier
# References

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: false
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{lineno}
  - \usepackage{etoc}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

options(digits = 2)
options(dplyr.summarise.inform = FALSE) # shut up about summarize

knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE,
        fig.height = 3.5
      )

knitr::knit_hooks$set(
  inline = function (x) {
    if (is.numeric(x)) x <- as.character(round(x, getOption("digits")))
    x <- gsub("-", "\u2212", x)
    paste(as.character(x), collapse = ", ")
  }
)

options(huxtable.long_minus = TRUE)
huxtable::set_default_properties(latex_float = "h!")

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] Note that equation
\eqref{eq:regression} does not control for many environmental and genetic
factors that could affect fertility, and as a result, $\beta$ is not an estimate
of the causal effect of a polygenic score on fertility. However, natural
selection is a correlational phenomenon: polygenic scores which correlate with
high fertility are being selected for, whatever the underlying causal
relationship.

@@

We also examine selection effects among respondents' parents, using
information on respondents' number of siblings. Effect sizes of polygenic
scores are highly correlated across the two generations. Median-splitting
respondents by year of birth, we find little evidence of change in effect sizes
among either the parents' or the respondents' generation. We check whether 
selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas
(Appendix Figure \@ref(fig:plot-siblings-townsend)). Lastly, the siblings data 
lets us check for a "quantity-quality tradeoff" between
number of children (of respondents' parents) and number of grandchildren. We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

@@

These results show that selection effects are weaker, absent, or even reversed
among some subgroups of the population. A possible explanation for this comes 
from the economic theory of fertility 
[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
increases in a person's wage affect their fertility via two opposing channels.
There is an *income effect* by which children become more affordable, like any
other good. There is also a *substitution effect*: since childrearing has a cost
in time, the opportunity cost of childrearing increases if one's market wage is
higher. The income effect leads higher earners to have more children. The
substitution effect leads them to have fewer.

@@

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is increasing and concave. We treat $N$ as continuous,
in line with the literature: this can be thought of as the expected number of
children among similar people. The marginal benefit of an extra child is
$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
marginal benefit is

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise.

@@

drake::loadd(res_all)
drake::loadd(res_age_flb)
drake::loadd(res_cor_educ)
drake::loadd(res_cor_income)

@@

dfr <- data.frame(
        PGS          = rep(score_names, 4),
        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
        cor          = rep(c(res_cor_income[,1], res_cor_educ[,1]), 2),
        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
        Controls     = rep(c("None", 
                        "Age at first live birth"), each = 66)
      )

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. The negative relationship
suggests that substitution effects dominate income effects, which fits the known
negative association between income and fertility
[@becker1960economic;@jones2006economic]. The correlations reverse when we
control for age at first live birth, suggesting that within AFLB categories, the
income effect dominates. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.)

```{r}
drake::loadd(res_mediation)

n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)

Second, we directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment is significantly different
from zero ($p < 0.05/33$).

@@

These results support our theory that natural selection on polygenic scores is
driven by their correlation with human capital. This need not imply
that a given polygenic score is causal of either fertility or human capital. Instead
scores may simply correlate with causally relevant environments, for example via
"genetic nurture" effects [@kong2018nature] or population stratification. To 
investigate the presence of these indirect genetic effects, we estimate equation
\@ref(eq:regression) among full siblings, controlling for a full set of sibling
group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
sample size, all within-sibling effects are insignificant after Bonferroni
correction. However, effect sizes are positively correlated with effect sizes
from the pooled model, and about 1/3 smaller (regressing within-sibling on
pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
with the ~50% decrease in heritability in within-sibling GWASs on age at first
birth and educational attainment [@Howe2021.03.05.433935]. We see these results
as providing tentative evidence that polygenic scores cause fertility, with effects
being at least partly driven by gene-environment correlations. We also reran
within-siblings regressions adding a control for education. Most effect sizes
barely change, suggesting that our measure of education does not in general
mediate differences in fertility among siblings. However, for educational
attainment polygenic scores, effect sizes are reduced by about 20%.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more
among low-income individuals. However, our effect sizes appear unchanged
over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
\@ref(tab:tbl-children-by-period)), during which
government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter $a$
in the model [cf. @jones2008fertility]. This theory would not explain why our
results are weaker at higher incomes and education levels.[^note] A third
alternative is that traits under selection are linked to externalizing behaviour
and risk-seeking. This might be partially captured by our risk-aversion
parameter $\sigma$, which affects fertility by changing the marginal utility of
income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of
children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
33 regressions). Given that, and the statistical power we get from our sample
size, we believe that the non-result is real: while risk attitude does predict
fertility in the sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced, or whether genetic
variants underlying other phenotypes will show a similar pattern to those
studied here. In addition, genetic effects on educational attainment have been
shown to be inflated in population-based samples as compared to within-family
designs, likely because of indirect genetic effects, gene-environment
correlations, and/or assortative mating [@lee2018gene; @selzam2019comparing;
@kong2018nature; @Howe2021.03.05.433935]. In short, it is probably too early to
tell whether modern natural selection has a substantively important effect on
population means of phenotypes under selection. Nevertheless, we note that selection
effects on our measured polygenic scores are still relatively small, even after
reweighting to account for ascertainment bias.

@@

If inequalities in polygenic scores are important for understanding social
structure and mobility [@belsky2018genetic;
@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
back-of-the-envelope calculation, using estimated heritability for educational
attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
income attributable to the "true" polygenic score for educational attainment as
`r r2_income_true_psea` among respondents. Natural selection would raise this
proportion to `r r2_income_true_psea_wt` among respondents' children,
counterfactually holding environmental conditions constant. Similarly, since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

Polygenic scores were chosen so as to cover a reasonably broad range of traits,
and based on the availability of a large and powerful GWAS which did not include
UK Biobank. Scores were computed by summing the alleles across ~1.3 million
genetic variants weighted by their effect sizes as estimated in 33 genome-wide
association studies (GWASs) that excluded UK Biobank. To control for population
stratification, we corrected the polygenic scores for 100 principal components
(PCs).

@@

\localtableofcontents
\clearpage

@@

The UK Biobank data contains information on respondents' number of
siblings, i.e. their parents' number of children. Since respondents' polygenic 
scores are equal in expectation
to the mean scores of their parents, we can use this to look at selection 
effects in the parents' generation. We estimate equation (\ref{eq:regression})
using *number of siblings* (including the respondent) as the dependent variable.
The parents' generation has an additional source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*.

Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
polygenic scores. For a clean comparison with the respondents' generation, we
rerun regressions on *number of children* excluding those with no children, and
show results in the figure. Selection effects are highly correlated across the
two generations, and most share the same sign. Absolute effect size estimates
are larger for the parents' generation. We treat this result cautiously, because
when we split respondents up by year of birth, we find few differences in effect
sizes between early- and late-born respondents, for either generation. In other
words, since estimated effect sizes change across "generations", but do not
change over time within either generation, the change may be due to remaining
ascertainment bias within the sample or other effects. In particular, effect
sizes in both generations may depend on polygenic scores' correlation with
childlessness, and we cannot estimate this for the parents' generation.

@@

The fact that childless people have such a strong effect on estimates makes it
hard to compare total effect sizes across generations. In particular, since the
parents' generation has a different distribution of numbers of children,
childless people may have had more or less effect in that generation. Another
issue is that we are estimating parents' polygenic scores by the
scores of their children. This introduces noise into our independent
variable, which might lead to errors-in-variables and bias coefficients towards
zero.

@@

drake::loadd(res_townsend_parents)
n_regs <- nrow(res_townsend_parents)

res_townsend_parents %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5
                  )

@@

n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)

@@

Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` scores, the
indirect effect on fertility via human capital is significantly greater than 0 
at $p = 0.05/33$.

@@

Table \@ref(tab:tbl-inequality-effects) shows correlations between children's
polygenic scores and household income (UKB data field 738). Column "With
selection" uses respondents' scores weighted by Age/Qualification, times number
of children. Column "Without selection" uses scores weighted by
Age/Qualification only, i.e. the counterfactual correlation if all respondents
had the same number of children.

@@

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ does not differ between the two sets of
weighted regressions.

The estimate above uses a pooled regression of income on EA3, which may include
confounding environmental variation such as genetic nurture. Within-siblings
regression would give a smaller estimate of the effect of EA3. However, this
would also affect the reliability of the regression [@griliches1986errors],
making it hard to back out estimates of the true $R^2$. Nevertheless, we
can calculate the proportional increase in $R^2$ due to natural selection, again 
so long as the reliability coefficient is unaffected by
the weighting scheme. Running within-siblings regressions of EA3 on income group
gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this procedure estimates the proportional change in $R^2_{income}$ due to natural selection as `r prop_wt_fe`.

@@

## Model proofs

@@

# References
